Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-19 Manuscript/manuscript.tex
New residence time section

The residence time of \ce{Ca^{2+}} in a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}.
\todo{This experimental estimate is actually for the residence time of Calcium
  ion at the individual PC lipid. This section is to be finished when the analysis details
and the results are here.}
The recent MD simulations indicate that the equilibration of \ce{Ca^{2+}}
binding to POPC bilayer may take hundreds of nanoseconds~\cite{javanainen17},
which could become an issue for the detailed binding studies with MD simulation
trajectories with the having lenght less than several microseconds \cite{catte16}.

%give a roughly estimate of the order
%of $1\,\mu\mathrm{s}$
%This is unfortunately not the case of many theoretical works,
%where the residence time cannot be estimated,
%or the simulation has not even decorrelated from it's 
%initial condition with a population of bound/unbound cations \cite{catte16}, 
%because it is limited by the simulation length, 
%even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
%Increasing time scales to the order of $\mu\mathrm{s}$ 
%doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used. 
To analyze the residence time in the ECC-lipid model we \todo{describe how the residence time analysis was done}.
Histograms of the residence times analyzed from the ECC-lipid simulation and from the CHARMM36 simulation data
\todo{specify the system, POPC?}
taken from the previous study \cite{javanainen17} (the simulation data was downloaded from Ref.~\citenum{zenodo.259376})
are shown in Fig.~S\ref{fig:hist_residence_times} in SI. 
\todo{JOE: Plot a histogram of residence times based on our simulations and simulation from \citenum{javanainen17} and put it into SI.
 }
%Based on Fig. S17 in that work, Ref.~\citenum{javanainen17},
%we took the simulation with the apparent highest fluctuations of the number of bound calcium,
%which can be obtained at \url{https://doi.org/10.5281/zenodo.259376}
%and anaylsed the residence times.
% simulation: (Matti C36 with CaClMK, 800ns total):

\todo{This paragraph should be clarified, but I cannot do it, because the residence time figure is not there yet.}
In the CHARMM36 simulation
approximately 20\% of bound residence times of \ce{Ca^{2+}} 
are limited by the length of the simulation, $800\,\mathrm{ns}$, 
and only less than
60\% of bound residence time is smaller than half of the simulation length , $400\,\mathrm{ns}$. 
Other simulations with CHARMM36 and Slipids models reported in the previous work show even longer residence times. 
The ECC-lipids model estimates 90\% of the residence times of calcium bound to a POPC membrane to be 
smaller than $60\,\mathrm{ns}$, % exactly $53\,\mathrm{ns}$
which is at least two orders of magnitude lower than for the CHARMM36.
In addition, the longest observed residence time in the ECC-lipid simulation is $120\,\mathrm{ns}$, 
which is well below the length of our simulations.

Interestingly, \ce{Na+} cations exhibit much more rapid exchange between membrane and solvent than calcium.
The residence time of sodium bound to a POPC membrane is
smaller than $2\,\mathrm{ns}$ for 90\% of the time it is bound to the membrane 
with the longest measured bound time being $13\,\mathrm{ns}$. 
Such an exchange between membrane and solvent 
is an order of magnitude faster than for calcium cations.

Such a finding changes the point of view on binding of calcium to PC membranes from
a very strong long-term stable binding with rare exchanges to 
a relatively frequent exchange of clacium cations 
in equilibrium between membrane and solvent. 
The reduced barrier for the Calcium binding is also visible in the density distributions
in Fig.~\ref{fig:cacl-dens}, where the more shallow minimum for the Calcuim
density is observed at the interface.

>>>>

Equilibration of \ce{Ca^{2+}} binding to POPC bilayer takes hundreds of
nanoseconds in MD simulations with CHARMM36 and Slipids force fields~\cite{javanainen17}, suggesting that
simulations with the length of several microseconds would be required
to study the binding phenomena in equilibrium. However, the density profiles of \ce{Ca^{2+}}
from the ECC-lipid simulation in Fig.~\ref{fig:cacl-dens} show a relatively shallow
minimum at the interface, suggesting a smaller barrier for the exchange of Calcium ions.
To quantify the ion binding exchange in lipid bilayer simulations, we calculated 
the residence times of ions within the distance of 0.3~nm from any oxygen atom
belonging to a lipid molecule in a bilayer.

The residence time histograms of \ce{Ca^{2+}}
in POPC bilayer are shown in Fig.~S\ref{fig:hist_residence_times} in SI 
from the ECC-lipid simulation with CaCl$_2$ concentration of 1~M
and from the CHARMM36 simulation with 450~mM concentration of
ECC-ions taken from the previous study \cite{javanainen17}
(the simulation data was downloaded from Ref.~\citenum{zenodo.259376}).
\todo{JOE: Plot a histogram of residence times based on our simulations and simulation from \citenum{javanainen17} and put it into SI.
}
\todo{It might be also better to do the analysis from ECC-lipid simulation with concentration closer
to the data in literature, i.e. 450mM.}
%Based on Fig. S17 in that work, Ref.~\citenum{javanainen17},
%we took the simulation with the apparent highest fluctuations of the number of bound calcium,
%which can be obtained at \url{https://doi.org/10.5281/zenodo.259376}
%and anaylsed the residence times.
% simulation: (Matti C36 with CaClMK, 800ns total):
Approximately 20\% of the \ce{Ca^{2+}} residence times in the CHARMM36 simulation are longer than
the lenght of the trajectory ($800\,\mathrm{ns}$),
and only less than 60\% of the bound residence times are shorter than the half
of the simulation length~(400~ns).
Even longer residence times are observed
in the other simulations with CHARMM36 and Slipids models reported in the previous work~\cite{javanainen17}.
At least order of magnitude faster exchange is observed in the
ECC-lipid model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are 
shorter than $60\,\mathrm{ns}$, % exactly $53\,\mathrm{ns}$
and the longest observed residence time in the ECC-lipid simulation is $120\,\mathrm{ns}$, 
which is well below the total length of the simulation (200~ns).
The exchange of \ce{Na+} ions to a POPC membrane is
an order of magnitude faster, giving 90\%
of the residence times smaller than~$2\,\mathrm{ns}$,
the longest residence time being~$13\,\mathrm{ns}$. 
All the results are in line with the experimental estimation that 
the residence time of \ce{Ca^{2+}} at the individual PC headgroup 
is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}.

%give a roughly estimate of the order
%of $1\,\mu\mathrm{s}$
%This is unfortunately not the case of many theoretical works,
%where the residence time cannot be estimated,
%or the simulation has not even decorrelated from it's 
%initial condition with a population of bound/unbound cations \cite{catte16}, 
%because it is limited by the simulation length, 
%even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
%Increasing time scales to the order of $\mu\mathrm{s}$ 
%doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used.

In conclusion, the results from ECC-lipid model suggest that the
exchange of bound Calcium occurs within $\sim$100ns timeframe,
which is much faster than previously reported~\cite{javanainen17}.
The sodium exhibit even faster exchange.
This suggests that simulations with the length of few hundred
nanoseconds are sufficient to simulate equilibrium ion binding
to PC lipid bilayers, when realistic force fields are used.
This has not been the case with the available lipid force fields,
which overestimate the binding streght of the cations \cite{javanainen17,catte16}.
%Such a finding changes the point of view on binding of calcium to PC membranes from
%a very strong long-term stable binding with rare exchanges to 
%a relatively frequent exchange of clacium cations 
%in equilibrium between membrane and solvent.
